---
title: "Billionaires EDA: Education & Industry"
author: "Jacob Edmonds"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
geometry: margin=1in
fontsize: 12pt
header-includes:
  - \usepackage{float}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(tidyverse)
})
billionaires <- read_csv("cleaning/billionaires_clean.csv")
```

## Research Question
In countries with a higher gross primary enrollment rate, do billionaires tend to have more diversified income sources or come from certain industries (e.g., tech vs. manufacturing)?

\newpage

```{r}
median_enrlmt <- median(billionaires$gross_primary_enrlmt_country, na.rm = TRUE)
```
## Industry Distribution by Enrollment
```{r, fig.width=8, fig.height=4.5, fig.pos='H'}
billionaires_enrlmt <- billionaires %>%
  filter(!is.na(gross_primary_enrlmt_country)) %>%
  mutate(enrlmt_bins = ifelse(gross_primary_enrlmt_country >= median_enrlmt, "High", "Low")) %>%
  group_by(enrlmt_bins, category) %>%
  summarize(count = n(), .groups = 'drop') %>%
  arrange(count)

ggplot(billionaires_enrlmt, aes(x = category, y = count, fill = enrlmt_bins)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap("enrlmt_bins") +
  coord_flip() +
  labs(
    title = "Industry Distribution of Billionaires by Enrollment Category",
    x = "Industry", 
    y = "Number of Billionaires") +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9),
    legend.position = "none"
  )
```
There are more billionaires with diversified income sources from countries with a primary enrollment above the global median than those from countries with primary enrollments below the global median. The distribution of billionaires from countries with higher gross primary enrollment rates is also different than that of the billionaires from countries with lower gross primary enrollment rates. For instance, there are many more billionaires from high-primary education countries in manufacturing, healthcare, and fashion and retail, and there are many more billionaires from low-primary education countries in finance & investments and technology. Countries were classified into high- and low-enrollment groups using the global median gross primary enrollment rate to ensure roughly balanced group sizes and to reduce sensitivity to extreme values.

\newpage

## Median Net Worth Analysis
```{r, fig.width=8, fig.height=4.5, fig.pos='H'}
billionaires_enrlmt_1 <- billionaires %>%
  filter(!is.na(gross_primary_enrlmt_country)) %>%
  mutate(enrlmt_bins = ifelse(gross_primary_enrlmt_country >= median_enrlmt, "High", "Low")) %>%
  group_by(category, enrlmt_bins) %>%
  summarise(
  median_net_worth = median(net_worth, na.rm = TRUE),
  .groups = "drop"
) %>%
  arrange(median_net_worth)

median_net_worth_total <- median(billionaires$net_worth)

ggplot(billionaires_enrlmt_1, aes(x = reorder(category,median_net_worth), y = median_net_worth)) + 
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap("enrlmt_bins") +
  geom_hline(yintercept = median_net_worth_total, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
   title = "Median Net Worth of Billionaires by Industry and Enrollment Category",
    x = "Industries",
    y = "Median Net Worth (Millions)"
  ) +
  theme_minimal() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 9),
    legend.position = "none"
  )

```
Looking at this distribution, the median net worth for billionaires from countries with higher gross primary enrollment rates seems to be less across most industries than the median net worth for billionaires from countries with lower gross primary enrollment rates in those same industries. For instance, billionaires from low-primary education countries earn, on average, more in logistics, finance & investments, and more, while billionaires from high-primary education countries earn more in industries like telecom. This suggests that the relationship between education levels and wealth accumulation may differ by industry structure rather than following a single global pattern.

### Limitations
This analysis relies on country-level education metrics, which may not reflect the individual educational backgrounds of billionaires. Additionally, dividing countries into high and low enrollment groups using the median simplifies variation in educational attainment across countries. Finally, net worth values represent a snapshot in time and may fluctuate substantially due to market conditions.